Title Pending

### 2019-02-25 manuscript/manuscript.tex

We first briefly review the formalism of Markov decision processes, which will provide the framework for our results. For complete treatments, see \cite{SuttonBarto1998, SuttonBarto2018} and \cite{bertsekas2005}.

@@

In the following sections, we show that our model is capable of recreating the results of several empirical studies of anxious cognition in sequential choice environments. We represent these experiments again using the formalism of MDPs. In all simulations, the values of environment states under the optimal policy were solved for directly through Q-value iteration \citep{SuttonBarto1998, SuttonBarto1998, bertsekas2005}. The details of this algorithm are presented in the appendix. All simulations were carried out with the python programming language and are publicly available on Github (https://github.com/szorowi1/SecretFunTimes).

@@

Next, we turn our attention to a related phenomenon observed in anxiety: a bias towards avoidance during approach-avoidance conflict \citep{aupperle2010}. In situations where the potential for reward and harm are correlated,
anxiety predicts the tendency to forego reward in favor of minimizing the risk of threat. As a consequence, individuals with anxiety may experience fewer rewards in their environment as compared to non-anxious individuals. For example, an individual with social anxiety may avoid a party for fear of social embarrassment at the expense of positive social interactions with friends. This is a primary mechanism by which clinical anxiety can disrupt everyday functioning.

A relationship between anxiety and biased decision making in approach-avoidance conflict is well-documented in the empirical literature. One common measure is the balloon analog risk task (BART) \citep{Lejuez2002}. The task is straightforward: a participant inflates a virtual balloon, one pump at a time. The more inflated the balloon, the more points the participant earns. With increasing pumps, however, comes the risk of the balloon popping and the loss of all earned points so far. Thus, there is a conflict: a participant can choose to pump the balloon more to earn more points, at the risk of the balloon popping; alternately, the participant can choose to end the trial earlier with fewer points earned. Previous studies have found that anxiety is correlated with earlier leaving times, or a greater tendency to avoid risk \cite{Maner2007, Giorgetta2012}.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/03_fid.png}}%
  }
  \par \textbf{Figure 3:} The Balloon Analog Risk Task (BART). Risk of the balloon popping increases with each pump (top row). An optimistic agent ($w=1$) continues to pump until the risk of losing all points earned outweighs the marginal value of one additional pump (second row). A pessimistic agent ($w=0.8$) is less confident that it will make the optimal decision in the future, and thus leaves earlier (third row). Leaving time becomes sooner with increasing pessimism ($w=0.6$) (bottom row). (Parameters: $\gamma = 1$.)
\end{figure}

This result can be accounted for by a loss of confidence in future reliability. A simulation of performance on the BART is shown in Figure 3. For an optimistic agent ($w=1$), the optimal policy is to continue pumping until the risk of popping is all but guaranteed. The optimal policy for pessimistic agents differ. Under a loss of confidence for future action, pessimistic agents cannot be certain they will make optimal choices in the future. As such, negative value propagates backwards resulting in earlier leaving times. The latency of leaving is dependent on the strength of the pessimistic belief.

Interestingly, these results are similar to recent findings regarding flight initiation distance and anxiety \citep{Mobbs2018, Mobbs2019}. Flight initiation distance refers to a measure from behavioral ecology that assesses the leaving time of a foraging animal at risk of predation. In classic flight initiation scenarios, an animal can continue to forage and exploit resources, but the longer it does so the greater the risk it will encounter a predator. Thus, flight initiation scenarios are similar in structure to approach-avoidance conflict and the BART. Recently, \cite{Mobbs2019} found that anxiety is correlated with earlier leaving times in virtual flight initiation tasks. Interestingly, the authors make a distinction between "fast-" and "slow-attacking" predators, and find that trait anxiety is only correlated only in behavioral response to the latter. The findings here draw upon an emerging literature of distinct neural systems mediating "cognitive" vs. "reactive" threat responses, where anxiety evidently impacts the deliberative, cognitive responses. Thus, it is important to note the present theory can account for anxious approach-avoidance behavior in some, but certainly not all, threatening scenarios.

These results may also be related to recent work on the computational mechanisms underlying behavioral inhibition, a temperament linked to anxiety \citep{bach2015, khemka2017}. \cite{bach2015} argues that delayed behavioral responses under threat are normative under the assumption that a pause in behavior (e.g. freezing) can be useful for surveying a threatening environment, avoiding detection by predators, and potentially waiting out a nearby threat. Behavioral inhibition, it is argued, is a useful Pavlovian heuristic that has evolved over time. The current models may predict similar results. Insofar that a loss of confidence in future action yields more negative estimates of action values, then the best action may often be brought closer in value to its alternatives. As such, a pessimistic agent may face increased decision conflict as compared to a non-anxious counterpart and therefore show increased response times. This is not to argue against this particular model of behavioral inhibition, but just to say that our account may yield similar behaviors.

\subsection{Relation to Depression}

As an extension of approach-avoidance conflict, we next want to discuss the relationship between anxiety and depression. Anxiety and depression are highly comorbid, with roughly 45\% of individuals with a lifetime depression diagnosis also diagnosed with at least one anxiety disorder \citep{kessler2015}. Given this, many have clinical theories assume that depression and anxiety are slightly variant manifestations of the same underlying mechanisms involving negative affect (for review, see \cite{jacobson2014}).

Where the two disorders can be distinguished is in relation to rewards and anhedonia; depressive disorders, but not anxiety disorders, are associated with disrupted reward processing. One interesting proposal in the literature is that at least certain types of anhedonic depression are preceded by anxiety \citep{alloy1990, moitra2008, jacobson2014}. Specifically, persistent avoidance has the long-run effect of reducing the likelihood of anxious individuals encountering positive events and activities in their everyday lives. As a consequence, anxious individuals come to expect an absence of reward in their environments and subsequently develop diminished reward motivation and behavioral activation. Recent large-scale epidemiological studies have provided  some empirical support for this claim, finding that anxiety symptoms and behavioral avoidance precede and predict the longitudinal onset of depression (\cite{mathew2011, jacobson2014, kessler2015}; though also see \cite{jacobson2017, plana2019}).

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/06_lh.png}}%
  }
  \par \textbf{Figure 4:} Progression from anxiety to depression. In this open environment, an agent starts opposite a reward (blue) with harm separating them (A). For an optimistic agent ($w=1$), positive value spreads from the reward to the starting tile (Panels B-D, left-to-right). Harm is mitigated through the belief of future self-efficacy. For a pessimistic agent ($w=0.6$), both positive and negative value spread (Panels E-G, left-to-right). Given the agent's closer proximity to harm than reward, the optimal policy for this agent is to stay at the initial state. (Parameters: $\gamma = 0.95$)
\end{figure}

This prediction is easily accommodated by the present model. Simulations of how this might come back is presented in Figure 4. In this environment, there is only one rewarding and harmful state. Just as in the open field environment, the world is deterministic and the harmful state poses no true threat to an agent. An optimistic agent ($w=1$), with a belief in its ability to maintain an optimistic policy, is unfazed by the presence of the harmful state and develops over time an overall positive representation of the value of the environment (Figure 4b-d).

In contrast, a pessimistic agent ($w=0.6$) cannot be certain of its ability to steer clear of future harm. As such, negative value spreads from the harmful state to its antecedent states (Figure 4e-g). Because in this environment risk and reward are correlated (i.e. the agent must approach the harmful state in order to reach the reward given its starting position), the consequence of negatively spreading value is that the optimal policy is to stay in the initial position. Thus, the pessimistic agent adopts a behavioral policy akin to depressive immobility. Moreover, reward is inaccessible to this agent such that its positive expectations of the environment are also diminished. In summary, a loss of confidence in future reliability can result in behavioral policies that resemble depressive-like states in environments with correlated risk and reward.

\subsection{Planning}

In this next section, we review the literature on aversive pruning. In large, multi-step sequential decision environments, the space of all possible sequences of choices grows exponentially larger with increasing sequence length. As such, it is infeasible for decision making agents to exhaustively explore all possible sequences select that which maximizes reward from the entire set. Heuristics for narrowing the search space then are not only plausible but beneficial for a decision making agent.

One such heuristic that has been proposed in the literature is aversive pruning \citep{Huys2012}. Aversive pruning is defined as a Pavlovian response to encountering a large loss in planning such that sequences involving large losses are discarded from further evaluation. An example of such a scenario is presented in Figure 5. The prediction in this decision tree environment is that an agent would discard the branch of the decision tree involving an immediate large negative loss, even though this branch objectively contains the reward maximizing (loss minimizing) sequence of choices.

\cite{Huys2012} (and later \cite{Lally2017}) tested this prediction in this particular choice environment. Participants received extensive training in this decision tree environment (first without outcomes and then later with rewards) in order to facilitate planning. In free-planning trials, many participants exhibited this Pavlovian bias, rejecting choice sequences which involved the large loss even when the sequence was objectively reward maximizing (loss minimizing). Importantly \cite{Huys2012} and \cite{Lally2017} found the degree of aversive pruning was correlated with depressive and anxiety symptoms, respectively.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/04_tree.png}}%
  }
  \par \textbf{Figure 5:} The decision tree environment from \cite{Huys2012} and \cite{Lally2017} (A). An optimistic agent ($w=1$) learns the optimal loss-minimizing policy through the initial large loss (B). A moderately pessimistic agent ($w=0.75$) will find both arms equally preferable insofar that it is less confident it will be able to execute optimal future choices (C). More pessimistic agents comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss (D).
\end{figure}

The aversive pruning principle makes tremendous sense and draws on a robust literature of heuristics and computational shortcuts for circumventing computationally intractable cognitive processes. It is important to note, however, that a control interpretation also makes similar predictions. Figure 5b-d shows the preferred routes of simulated agents in the same environments under increasingly pessimistic expectations of future actions. As can be observed, pessimistic agents are similarly likely to avoid the branch with the large loss, though for different reasons. Under the pessimistic case, the agent is increasingly less confident that the large gains later in sequence will be realized; as such, the agent is less confident that the initial large loss will be offset. Consequently, pessimistic agents would prefer the branch with smaller losses (even if this is objectively suboptimal). This prediction is similarly consistent with the finding that avoidance of the initial large loss is correlated with anxiety \cite{Lally2017}.

%% An interpretation like this would be consistent with reduced beliefs in one's own competency in such an environment. This is not strictly surprising. Planning multi-choice sequences even in relatively simple environments like those used in these experiments still ostensibly require working memory ability to keep in mind the possible consequences along each branch. Insofar that working memory ability is disrupted in anxiety disorders \citep{Moran2016}, a pessimistic belief may be justified and possibly optimal.

It is worth noting that these two alternate accounts can be disentangled through slight manipulations of the choice environment presented above. Aversive pruning and loss of confidence make similar predictions when a large loss begins a sequence, but not necessarily when a large potential loss comes later in a sequence. In the loss of confidence view, the potential for large loss late in a sequence negatively impacts the estimated value of actions early in the sequence; this is not true for aversive pruning. Identifying such choice sequences in the decision tree task may help identify whether this sort of inference is present in individuals with anxiety performing the task.

@@

In this final section, we propose based on the present theory that anxiety should be inversely related to the free choice bias. If an agent is confident in their future decision making abilities, then it follows that situations in which an individual is able to make a future choice should be preferable to situations in which an individual is not. At worst, an agent is no better off than if they had not made a choice; at best, an agent is better able to steer themselves toward desirable outcomes. Such is the logic that underlies an emerging literature suggesting that choice is inherently valuable \citep{Leotti2010}.

This free choice premium has been observed across several behavioral experiments \citep{Suzuki1997, Leotti2011, Leotti2014, Cockburn2014} using a variety of decision paradigms. For the present purposes, we will describe the experiments presented in \citep{Leotti2011, Leotti2014}. In these experiments, participants complete a two-stage decision making task (Figure 6). In the first stage, participants are make a choice between two cues: one leading to free choice and the other leading to a fixed choice. In the second stage, participants are able to choose between a second set of cues (free choice) or are forced to choose a cue (fixed choice). Unbeknownst to participants, all cues yield reward following the same outcome distribution; thus, no cue is advantageous in the long-run. The free choice premium is measured as a preference for the cue leading to free choice. \cite{Leotti2011, Leotti2014} find participants still prefer the free choice cue, despite this conferring no objective benefit to participants.

Insofar that the free choice premimum is contingent on an agent's belief in its future ability, it stands to reason that this free choice premium should be absent in individuals with anxiety and a loss of confidence. Simulations of this effect are presented in Figure 6. Simulated agents complete an experiment similar to that used in \cite{Leotti2011, Leotti2014}. In contrast to optimistic agents, the value of the worst action in the free choice arm back-propagated to the free-choice cue in pessimistic agents. With increasing pessimism then, the free choice option is less preferable than the fixed choice arm. As a result, pessimistic learners, on average, do not exhibit a bias towards the free choice arm. To our knowledge this prediction has not been empirically tested.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/05_choice.png}}%
  }
  \par \textbf{Figure 6:} Schematic of the free choice bias tasks used in \cite{Leotti2011, Leotti2014} (A). The free choice bias is diminished in increasingly pessimistic agents (B).
\end{figure}

@@

\section{Appendix}

\begin{algorithm}
  \caption{Value Iteration}

  \State Algorithm parameter: a small threshold $\theta > 0$ determining accuracy of estimation
  \State Initialize $V(s)$, for all $s \in S$ arbitrarily, except that $V(terminal) = 0$
  \State
  \While{$\Delta > \theta$}
    \State $\Delta \leftarrow 0$
    \Loop \ for each $s \in S$
      \State $v \leftarrow V(s)$
      \State $ V(s) = \max_a \sum_{s',r} p(s',r|s,a) \left[ r + \gamma V(s') \right] $
      \State $\Delta \leftarrow \max(\Delta, |v - V(s)|)$
    \EndLoop
  \EndWhile

\end{algorithm}

>>>>

We first briefly review the formalism of Markov decision processes, which will provide the framework for our results. For complete treatments, see \cite{SuttonBarto2018} and \cite{bertsekas2005}.

@@

In the following sections, we show that our model can account for the majority of extant empirical studies of anxious cognition in sequential choice environments. We represent these experiments again using the formalism of MDPs. In all simulations, the values of environment states under the optimal policy were solved for directly through Q-value iteration \citep{bertsekas2005, SuttonBarto2018}. All simulations were implemented using the python programming language and are publicly available on Github (https://github.com/szorowi1/SecretFunTimes).

@@

In the broadest of terms, the motivations for action can be characterized with respect to two fundamental drives: the drive to approach rewards and the drive to avoid harm \citep{corr2013}. Though usually aligned, these motivations are necessarily antithetical wherever reward and risk of harm are correlated. In approach-avoidance conflict, an agent must integrate information regarding potential gains and losses in order to decide whether to perform an approach or avoidance action. Unsurprisingly, approach-avoidance conflict processing been suggested as integral to understanding the anxiety disorders \citep{aupperle2010} insofar that many of the disruptions they cause to everyday functioning (e.g. avoiding social obligations for fear of possible social humiliation) can be recast as biased arbitration of approach-avoidance conflict.

Many experimental protocols exist for studying approach-avoidance conflict behavior. One popular paradigm is the balloon analog risk task (BART) \citep{Lejuez2002}. In the BART, participants are offered the opportunity to win money by pumping virtual balloons. With each successful pump, the balloon incrementally inflates and money is added to the participant's till. Following one too many pumps, however, the balloon bursts and all money earned on that trial is lost. At any point during a trial, a participant may also choose to cash out, thereby banking the money earned up to that point and ending the trial. Thus, the BART pits approach and avoidance motivations against each other: with each pump, the reward earned increases but so too does the risk of losing it all. Importantly, previous studies have found that anxiety is correlated with fewer pumps and earlier cash-out times \citep{Maner2007, ramirez2015}.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/02_bart.png}}%
  }
  \par \textbf{Figure 2:} The Balloon Analog Risk Task (BART). The risk of balloon burst (point loss) increases with each pump (A). The optimal policy (number of pumps) under increasingly pessimistic learning is presented for low risk (B) and high risk (C) balloons. The optimistic agent ($w=1$) learns a policy reflecting the objective environmental risk. The moderate ($w=0.6$) and strongly ($w=0.2$) pessimistic agents learn to cash-out earlier, as is observed in anxious individuals. (Parameters: $\gamma = 1.0$)
\end{figure}

This empirical result is easily accounted for by our model. A simulation of performance on the BART is presented in Figure 2. For an optimistic agent ($w=1.0$), the optimal policy accurately reflects the statistics of the task. The optimal agent continues pumping the balloon until the marginal gain from one additional pump no longer exceeds the likelihood of of the balloon bursting. Thus, the optimistic agent continues pumping until reaching the (unobserved) mean number of pumps before the balloon bursts. The same is not true for the pessimistic agents. The moderately ($w=0.6$) and strongly ($w=0.2$) pessimistic agents exhibit behavioral policies with earlier cash-out times. The reason for this is that anticipated future harmful actions are able to exert influence on earlier states; risky pumps resulting in loss are more salient to these agents earlier in the trial. As such, the pessimistic agents perform similarly to what is observed empirically in anxious individuals.

It is worth noting that these results are similar to recent findings regarding flight initiation distance (FID) and anxiety \citep{Mobbs2018, Mobbs2019}. FID refers to a concept from behavioral ecology that measures the leaving time of a foraging animal at risk of predation. In the prototypical FID scenario, an animal can choose either to continue exploiting resources (i.e. accrue reward) or flee an approaching predator. Similar to the BART, the longer an animal chooses to exploit the greater reward and risk it experiences. In a recent study, \cite{Mobbs2019} investigated the relationship between sub-clinical levels of anxiety and leaving times during a virtual FID task involving both fast- and slow-acting predators. Interestingly, sub-clinical anxiety did not correlate with leaving times for fast predators, but was correlated with earlier leaving times for slow predators. (Sub-clinical anxiety also correlated with BOLD activity in brain structures including the ventral hippocampus, medial prefrontal cortex, amygdala and insula.) The authors use these results to support an argument distinguishing between "reactive" responses associated with immediate threat and "cognitive" responses associated with impending danger. Similar to the BART results, our model can readily account for the "cognitive"-type threat response reported in \cite{Mobbs2019}; importantly though, these findings suggest that our model may not be applicable to all types of threat processing, such as those with little reliance on future expectations (such as in immediate threat).

Similarly related are recent studies on behavioral inhibition, a behavioral style characterized by a temporal delay to initiate approach behavior associated with the development of anxiety disorders \citep{bach2015, khemka2017}. \cite{bach2015} argues that behaviorally inhibited responses (e.g. slowed reaction times under threat) reflect a normative Pavlovian heuristic, developed over evolutionary time, that is instrumental in surveying the environment, avoiding detection by predators, and possibly even waiting out nearby threat. Evidence for this heuristic comes from a sequential approach-avoidance conflict task, wherein a participant may choose to exit a safe zone to collect valuable tokens adjacent to a "sleeping" virtual predator. With each token collected, the probability of the predator "waking up", unbeknownst to the participant, increases. Similar to the BART, if the predator catches the participant, all tokens accrued during that trial are lost. Importantly, \cite{bach2015} (and later \cite{khemka2017}) find that response times for choosing to leave the safe zone increase as a function of predation threat, even though response times have no bearing on whether the predator "awakes", in line with the interpretation of behavioral inhibition as Pavlovian heuristic.

Our model can offer an alternative interpretation. As observed above, one consequence of uncertainty about avoiding harmful future actions is that antecedent actions become associated with threat. In the context of the sequential approach-avoidance task, our pessimistic bias should therefore reduce the value of the choice to leave the safe zone to collect tokens, making it closer in value to the choice to stay and forego future point collection. Insofar that choices between similarly valued options result in prolonged response times by virtue of increased decision conflict, then we should predict that anxious individuals would exhibit, in addition to earlier leaving times, slower responses in sequential approach-avoidance task independent of a Pavlovian heuristic. Further work is necessary to disentangle these two interpretations.

\subsection{Anxiety-to-Depression Transition}

As will become clear, the previous section concerning approach-avoidance conflict is a natural interlude to discussing the relationship between anxiety and depression. Anxiety and depression are highly comorbid, with roughly 45\% of individuals with a lifetime depression diagnosis also diagnosed with at least one anxiety disorder \citep{kessler2015}. Because of their frequent co-occurrence, many clinical theories treat anxiety and depressive disorders as variant manifestations of the same latent mechanisms causing disruption to negative affective processes (for review, see \cite{jacobson2014}). Where the two families of disorders can be distinguished is in their relation to positive affect and reward processing. Some have argued that whereas both anxiety and depression both involve exaggerated threat appraisals and pessimistic future expectations, only depression involves the certain belief about the absence of future rewards.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/03_lh.png}}%
  }
  \par \textbf{Figure 3:} Progression from anxiety to depression. In this open environment, an agent starts opposite a reward (blue) with harm separating them (A). For an optimistic agent ($w=1$), positive value spreads from the reward to the starting tile (Panels B-D, left-to-right). Harm is mitigated through the belief of future self-efficacy. For a pessimistic agent ($w=0.6$), both positive and negative value spread (Panels E-G, left-to-right). Given the agent's closer proximity to harm than reward, the optimal policy for this agent is to stay at the initial state. (Parameters: $\gamma = 0.95$)
\end{figure}

One interesting proposal in the clinical literature attempting to explain this divergence appeals to longitudinal association between anxiety disorders and depression. Specifically, it has been proposed that the onset of at least certain types of anhedonic depression are preceded first by the onset of an anxiety disorder \citep{alloy1990, moitra2008, jacobson2014}. The idea according to this view is that because anxiety is associated with a bias towards avoidance behavior, then anxious individuals should be more likely to forego reward and other positive events in their everyday lives. Over time, it is argued, this should translate to a stable belief that, in addition to threat being unavoidable, reward is unobtainable thereby resulting in reduced behavioral activation and increased motor inhibition. Recent large-scale epidemiological studies have provided some empirical support for this claim, finding that anxiety symptoms often precede depression (\cite{mathew2011, jacobson2014, kessler2015}; though also see \cite{jacobson2017, plana2019}). Interestingly, one study event found that anxious avoidance behaviors were predictive future depression onset over a decade later \citep{jacobson2014}.

Though currently speculative, this prediction is accommodated by our model. Simulations of how this might come about is presented in Figure 3. We assume an open grid world, similar to that presented in Figure 1, except that rewards and harms are now (spatially) correlated. Figure 3a shows that, in order to reach the reward, an agent must approach (but need not encounter) danger. As above, the world is deterministic such that the presence of the harmful state poses no objective danger to the agent. The learning progression of an optimistic agent ($w=1$) reflects this (Figure 3b-d). An optimistic agent believes it will be able to act in accordance to its reward-maximizing preferences in the future; as such, it does not worry that it may inadvertently encounter danger in the future. As a consequence, only reward opportunity propagates from the source to its antecedent states whereas harm is effectively sequestered. The optimistic agent thereby learns a path from the start to reward in spite of the presence of a dangerous state.

In contrast, a pessimistic agent ($w=0.6$) is uncertain about its ability to effectively avoid future danger. As a consequence, reward and threat values propagate from their respective sources to their antecedent states (Figure 3e-g). Because reward and threat are (spatially) correlated in this environment, however, subjective threat increases with proximity to reward. As such, the pessimistic agent learns a behavioral policy distinct from the optimistic agent. Whereas the optimistic agent learns a path which circumnavigates danger, the pessimistic agent learns that any step toward reward is also a step towards danger. Consequently, the pessimistic agent learns to not move. In effect, the pessimistic agent exhibits a form of motor inhibition and immobility like that observed in depression. Moreover, as a function of its learned behavioral policy, the agent should not expect reward from the environment. In summary, our model can accommodate previous clinical theories about the transition from anxiety to depression; specifically, our model predicts that when reward and danger are correlated, pessimistic expectations and anxious avoidance can result over time in depression-like motor inhibition.

\subsection{Planning \& Aversive Pruning}

In large sequential choice environments, the total number of possible choice sequences grows exponentially with sequence length. As such, it is infeasible (if not impossible) for an agent to exhaustively evaluate all possible sequences of choices in order to carry out the one which maximizes reward. It is plausible then that rational agents may develop useful heuristics for winnowing out likely sub-optimal choice sequences. One such heuristic that has been proposed is aversive pruning \citep{Huys2012}, which is characterized as a Pavlovian response to losses such that choice sequences involving large losses are discarded from further evaluation during planning.

\cite{Huys2012} (and later \cite{Lally2017}) sought evidence for aversive pruning in the decision tree environment presented in Figure 4a. Participants in these experiments received extensive training in this environment such that participants reportedly knew well both the transition and reward structure of the tree. During testing, participants completed a number of free-planning trials in which participants started at various states within the tree and selected paths along the tree so as to maximize reward. \cite{Huys2012} and \cite{Lally2017} predicted that participants starting at the top of the tree (as depicted in Figure 4a) would be more likely to discard the left branch as it involved an immediate large loss and would likely be discarded through aversive pruning (and despite the fact that it was the parent path to the reward-maximizing choice sequence). Both \cite{Huys2012} and \cite{Lally2017} reported that many participants exhibited this pattern of behavior in evidence of an aversive pruning heuristic. Most importantly for present purposes, \cite{Huys2012} and \cite{Lally2017} found the degree of aversive pruning was correlated with depressive and anxiety symptoms, respectively.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/04_tree.png}}%
  }
  \par \textbf{Figure 4:} The decision tree environment from \cite{Huys2012} and \cite{Lally2017} (A). An optimistic agent ($w=1$) learns the optimal loss-minimizing policy through the initial large loss (B). A moderately pessimistic agent ($w=0.75$) will find both arms equally preferable insofar that it is less confident it will be able to execute optimal future choices (C). More pessimistic agents comes to prefer the branch without the large loss so as to avoid being unable to recoup the large initial loss (D).
\end{figure}

The aversive pruning heuristic is an intuitive and sensible strategy for circumventing exhaustive search, which draws upon a rich literature of Pavlovian biases. We note, however, that our model makes similar predictions in the decision tree environment employed by \cite{Huys2012} and \cite{Lally2017}. Assuming exhaustive training, the optimal policy for an optimistic agent ($w=1.0$) beginning at the top of the tree is presented in Figure 4b. Note the optimistic agent traverses through the initial large loss in order to complete the sequence that maximizes reward (minimizes loss). Different patterns of behavior are exhibited by moderate ($w=0.75$, Figure 4c) and strongly ($w=0.50$, Figure 4d) pessimistic agents. For these agents, the possibility of inadvertently choosing a future sub-optimal path, and accruing a large loss, are non-negligible. As such, preference for the objectively sub-optimal path not involving the large loss becomes increasingly preferred. This preference is reminiscent of what is observed for the subclinically depressed and anxious participants in the original studies, suggesting that our model may also describe anxious decision making in this task. This is not implausible insofar that working memory ability is often impaired in anxiety disorders \citep{Moran2016}, and pessimistic expectations about future choice may be warranted as such.

We note that this reinterpretation of results is not meant to replace the aversive pruning heuristic. Our model makes no strong claims about mechanistic interpretations, and thus does not address the issue of computational intractability whereas the aversive pruning heuristic does. We are simply noting that they make similar behavior predictions for anxious decision making in the decision tree environment used thus far. Interestingly, these accounts could be disentangled through a minor manipulation of the choice environment presented above. One would simply need to move the large loss to later in the tree. Under the aversive pruning view, large losses only impact choice behavior when encountered in sequence. Under pessimistic sequential choice inference, however, future large losses impact antecedent states, making their selection less likely. Thus, a slight modification to the decision tree structure presented in Figure 4a might be used to disentangle which of the two theories better accounts for anxious decision making in the task.

@@

In this final section, we offer a novel prediction regarding the free choice premium. If an agent assumes they will be able to act in accordance to their (reward-maximizing) preferences in the future, then it follows that choices that enable an individual to make a future choice should be preferable to situations in which an individual is not. At worst, an agent is no better off than if they had not made a choice; at best, an agent is better able to steer themselves towards a desired outcome. Such is the logic that underlies an emerging literature suggesting that choice is inherently valuable \citep{Leotti2010}.

A free choice premium (i.e. choice is inherently valuable, all else being equal) has been observed across several behavioral experiments \citep{Suzuki1997, Leotti2011, Leotti2014, Cockburn2014} using several different experimental protocols. For present purposes, we will describe the protocol used by \cite{Leotti2011, Leotti2014}. Here, participants complete a two-stage decision making task (Figure 5a). In the first stage, participants choose between two cues: one yielding a future free choice and the other yielding a future fixed choice. In the second stage, participants either make a second choice (free choice) or have a choice made for them (fixed choice). Unbeknownst to participants, all choices in the second stage yield a reward drawn identical reward distributions; thus, no choice is actually advantageous in the long-run. The free choice premium is measured as a preference for the cue leading to free choice. \cite{Leotti2011, Leotti2014} find that participants prefer the cue leading to future choice, despite this conferring no objective benefit to participants.

We hypothesize that the free choice premium is contingent on an agent's positive expectations about its future actions. If this is true, it stands to reason that the free choice premium should be absent in anxious individuals. We present simulations of this effect in Figure 5b using a virtual experiment similar to that used in \cite{Leotti2011, Leotti2014}. As expected, optimistic agents ($w=1.0$) exhibit the free choice premium, measured as a preference for the cue leading to future choice. In contrast, moderately ($w=0.5$) and strongly ($w=0.0$) pessimistic agents exhibit a diminished free choice bias. In these agents, the value of worst action in the free choice is able to propagate backwards to the free choice action in stage 1. As such, the free choice confers no subjective benefit to the simulated agents and thus the free choice premium disappears. We plan on explicitly testing this hypothesis in subclinical and clinical anxiety in the near future.

\begin{figure}
  \centerline{%
    \resizebox{1.0\textwidth}{!}{\includegraphics[trim={0 0 0 0},clip]{../figures/05_choice.png}}%
  }
  \par \textbf{Figure 5:} Schematic of the free choice bias tasks used in \cite{Leotti2011, Leotti2014} (A). The free choice bias is diminished in increasingly pessimistic agents (B).
\end{figure}

@@
